%% ASPLOS 2026 submission template
%% Class:  acmart (sigplan proceedings)
%% Per ASPLOS'26 CFP: \documentclass[sigplan,nonacm]{acmart}
%% Body text 10pt. Page numbers enabled via \settopmatter{printfolios=true}.
%% Bibliography: ACM-Reference-Format.
%% This is the Project Checkpoint document (2--3 pages).

\documentclass[sigplan,nonacm,review]{acmart}

\settopmatter{printfolios=true,printccs=false,printacmref=false}

%% ---- Packages -----------------------------------------------------------
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{balance}

%% Give TeX more wiggle room to avoid overfull hbox in narrow columns
%% (the tt-metal paths and underscored identifiers are long).
\emergencystretch=5em
\hbadness=10000
\tolerance=2000

%% ---- Metadata (stripped by nonacm, but kept for completeness) -----------
\acmConference[ASPLOS '26]{31st International Conference on Architectural
  Support for Programming Languages and Operating Systems}{March 22--26,
  2026}{Pittsburgh, PA, USA}

%% ---- Title --------------------------------------------------------------
\title{Optimizing Qwen2.5-Coder Inference on Tenstorrent Blackhole:\\
       Project Checkpoint}

\author{Rassul Magauin}
\author{Assylzhan Khamiyev}
\author{Mohammed Rashed Ali Yahmoor Alshehhi}
\author{Sultan Akimaliyev}
\author{Hamad Khalifa Alyahyaee}
\affiliation{%
  \institution{Mohamed bin Zayed University of Artificial Intelligence}
  \city{Abu Dhabi}\country{UAE}}

\renewcommand{\shortauthors}{Magauin et al.}

%% ---- Abstract -----------------------------------------------------------
\begin{abstract}
This checkpoint describes our progress on optimizing Qwen2.5-Coder
inference for the Tenstorrent Blackhole p150 using the tt-metal SDK. We
began by profiling the existing \texttt{tt\_transformers} port of
Qwen2.5 with the stock Blackhole tuning recipe on a four-device
tensor-parallel mesh. Two things surprised us. First, the single
largest kernel in the profile is TopK at 45--60\% of per-token device
kernel time, but it runs on only one of the 130 Tensix cores, so it is
really a sampling-stage mapping bug rather than a layer bottleneck.
Second, the fused SDPA kernel is already very cheap (under 1.5\%),
which tells us attention is not where we should be spending effort.
Once TopK is set aside, the MLP feed-forward block is the real hot
path at about 31\% of kernel time, with the gate/up and down
projections doing most of the work. Based on this, we plan to
parallelize TopK, try dropping the MLP matmuls from HiFi4 to HiFi2
math fidelity, and fuse the gate and up projections into one wide
matmul. We report baseline numbers, ranked targets, and the roadblocks
we ran into along the way.
\end{abstract}

\begin{document}
\maketitle

%% =========================================================================
\section{Introduction and Goals}
%% =========================================================================
Large language models for code, such as Qwen2.5-Coder~\cite{qwen25coder},
are increasingly being targeted at non-GPU accelerators, and our course
project picks one such pairing: Qwen2.5-Coder on Tenstorrent's Blackhole
p150. Blackhole is a RISC-V-based tensor processor with 130 worker
Tensix cores. Each core has a matrix unit (FPU), a vector unit (SFPU),
five Baby RISC-V CPUs, and 1.5~MB of L1 SRAM, all wired together with a
2D NoC~\cite{ttmetal}. Tenstorrent's programming model, \texttt{tt-metal},
lets the developer reach all the way down to kernels, circular buffers,
and NoC scheduling. That gives us a lot of flexibility, and also a lot
of room to make things slower than they should be.

\paragraph{Direction.} Our original plan was to port Qwen2.5-Coder to
Blackhole from scratch. After talking with our TA we switched to
Direction~B (optimization), because the in-tree
\texttt{tt\_transformers} framework already supports the Qwen2.5
family. The new goal is to profile, rank, and then fix whichever
kernels are actually slow on Blackhole.

\paragraph{What this checkpoint covers.} We set up a profiling pipeline
for \texttt{tt\_transformers} decode on a four-device Blackhole mesh,
using the Tracy profiler together with the \texttt{tt-perf-report}
utility our TA recommended. We then used that pipeline to produce a
per-operator breakdown of Qwen2.5-Coder-7B-Instruct decode against the
stock tuned recipe, and we used the breakdown to decide what to work on
next. Three optimizations came out of this analysis, and we describe
them in Section~\ref{sec:opts}.

%% =========================================================================
\section{Background}
%% =========================================================================
\paragraph{Blackhole p150.}
The compute grid on Blackhole is $8{\times}10$ (130 worker cores visible
to \texttt{tt-metal}), compared to Wormhole's $8{\times}8$. Each core
has 1.5~MB of L1 SRAM, and there are eight DRAM banks split across the
chip. The in-tree \texttt{tt\_transformers} code has several
Blackhole-specific branches gated on \texttt{is\_blackhole()}. The most
visible ones are a 512-token prefill-length cutoff (Wormhole uses 1024)
and a Blackhole-only DRAM weight prefetcher that is currently wired up
only for Llama-3.1-8B.

\paragraph{Qwen2.5-Coder-7B-Instruct.}
We checked the Hugging Face configs and found that Qwen2.5-Coder-7B
shares its base transformer with Qwen2.5-7B-Instruct. Both have 28
layers, hidden size 3584, MLP intermediate size 18944, 28 query heads
with 4 grouped key/value heads, and \texttt{rope\_theta} $=10^{6}$.
This matters for us because \texttt{tt-metal} already ships a tuned
performance decoder config for Qwen2.5-7B, and we can reuse it
directly.

\paragraph{Fusion in \texttt{tt-metal}.}
\texttt{tt-metal} does not fuse kernels by writing one monolithic
kernel. A reader kernel streams tiles into L1 circular buffers, a
compute kernel drains them through the FPU/SFPU, and a writer kernel
streams the results out. ``Fusion'' is really a property of how those
buffers are wired up. We lean on five patterns that already exist in
the repo: fused SDPA, matmul${+}$bias${+}$activation, distributed
RMSNorm, asynchronous AllGather${+}$matmul, and the paged KV-cache
update.

%% =========================================================================
\section{Methodology}
%% =========================================================================
\paragraph{Hardware.} One machine with four Blackhole p150 devices at
\texttt{/dev/tenstorrent/\{0..3\}}, running a \texttt{tt-metal} build
with \texttt{ENABLE\_TRACY=ON}.

\paragraph{Workload.} We drive the canonical
\texttt{simple\_text\_demo.py} benchmark from
\texttt{tt\_transformers} in decode mode: batch~1, 1024-token context,
10 transformer layers, paged attention on, and the stock
performance decoder config. The \texttt{blackhole-4} pytest fixture
picks a 4-way tensor-parallel mesh for us. Weights are local
\texttt{bf16} safetensors downloaded from Hugging Face.

\paragraph{Profiling.} We wrap the pytest invocation in
\texttt{python -m tracy} with the \texttt{device\_kernel\_duration}
analysis turned on. That writes a per-op CSV under the profiler
reports directory (\texttt{ops\_perf\_results\_*.csv}), which for our
10-layer run contained 4{,}916 rows across two trace-replay sessions
(compile phase excluded). We process the CSV two ways. First, a small
pandas script for per-op aggregates. Second,
\texttt{tt-perf-report}~v1.2.2~\cite{ttperfreport}, which gives us a
stacked per-op-class view and, more usefully for us, math-fidelity
annotations.

%% =========================================================================
\section{Baseline Results}
%% =========================================================================
Table~\ref{tab:baseline} shows the per-iteration device-0 kernel time
for Qwen2.5-Coder-7B-Instruct decode on the four-way Blackhole mesh,
taken from steady-state trace replay with the compile phase excluded.
Per-iteration device kernel time is 12.1~ms for the 7B-Instruct run and
10.4~ms for the Coder-7B rerun. The two runs have bit-identical op
counts and shapes, and the TopK kernel time matches to the nanosecond,
so the gap is almost certainly compile-cache warm-up between two
back-to-back runs. We will redo this with a cold cache before the final
report.

\begin{table}[t]
\centering
\small
\caption{Per-iteration device-0 kernel time, Qwen2.5-Coder-7B-Instruct
  decode, 4-way Blackhole TP mesh, 10 layers, stock tuned recipe.
  TopK is a sampling-stage op, not a transformer-layer op.}
\label{tab:baseline}
\begin{tabular}{lrr}
\toprule
Operator / group & Time (ms) & \% of total \\
\midrule
TopK (sampling, 1-core)          & 5.48 & 45.3 \\
MLP gate${+}$up projections      & 2.41 & 19.9 \\
MLP down projection              & 1.40 & 11.5 \\
Attention QKV projections        & 1.03 &  8.5 \\
CCL (AllGather ${+}$ ReduceScatter)& 0.69 &  5.7 \\
Layout/tiling/eltwise glue       & 0.46 &  3.8 \\
Attention output projection      & 0.28 &  2.3 \\
Fused SDPA decode ${+}$ RoPE     & 0.15 &  1.3 \\
RMSNorm                          & 0.13 &  1.1 \\
Other                            & 0.08 &  0.6 \\
\midrule
\textbf{Total per iter.}         & \textbf{12.11} & \textbf{100.0} \\
\bottomrule
\end{tabular}
\end{table}

%% =========================================================================
\section{Bottleneck Analysis}
%% =========================================================================
\paragraph{TopK is a sampling-stage bug, not a layer bottleneck.}
\texttt{TopKDeviceOperation} takes 45--60\% of per-token device kernel
time, with input shape $[1,1,32,38016]$ and $k{=}32$. What makes this
strange is that the kernel only runs on a single Tensix core, out of
the 130 that Blackhole exposes. So it is not really a transformer-layer
issue at all. It is a sampling-stage mapping bug, and an existing
multi-core reduction template should remove most of it. Of the three
things we are planning to try, this one looks the easiest and carries
the least risk, so we want to do it first.

\paragraph{The MLP block is where the real time goes.}
Once TopK is set aside, the MLP feed-forward block is the hot path:
about 31\% of kernel time per iteration, or roughly 3.81~ms. It splits
into the gate and up projections ($3584 \!\to\! 5120$, 20 calls per
iteration) and the down projection ($5120 \!\to\! 3584$). The single
slowest op in the whole profile is the down projection at 139.8~$\mu$s
peak. \texttt{tt-metal}'s model config file shows these matmuls running
at HiFi4 math fidelity on BF16/BFP8 operands.

\paragraph{Attention is already fused and cheap.}
The fused SDPA-decode kernel, the head reshapes, RoPE, and the
KV-cache update together add up to only 1.3\% of per-iter kernel time.
We take this as a good sign: the flash-attention-style CB choreography
already shipped under \texttt{ttnn/.../transformer/sdpa/} is doing its
job. We do not plan to touch attention.

\paragraph{Tensor-parallel overhead is small but present.}
AllGather plus ReduceScatter together add up to 5.7\%. This is headroom
we could claim later, but only after the MLP wins are in.

%% =========================================================================
\section{Planned Optimizations}
\label{sec:opts}
%% =========================================================================

\paragraph{Parallelize or fuse TopK.} Moving
\texttt{TopKDeviceOperation} off a single core onto a multi-core
reduction, or fusing it straight into
\texttt{SamplingDeviceOperation}, should give us most of the 45--60\%
TopK slice back. The effort looks low: this is mostly a dispatch and
sharding question, not a new kernel. Because sampling happens
once per generated token, any speedup here applies to decode regardless
of how many layers the model has.

\paragraph{HiFi4 $\to$ HiFi2 for the MLP matmuls.} Blackhole's FPU has a
HiFi2 mode that trades the lowest activation bit for about a
2$\times$ matmul throughput. \texttt{tt-perf-report} flagged this as the
biggest single-line change available against the stock recipe. We
expect it to cut the MLP slice by up to half, which is around
1.9~ms/iter, or roughly 15\% of the TopK-excluded kernel budget. The
change is one field in the performance decoder config. The main risk
is an accuracy regression, so we will run the in-tree
\texttt{ci-token-matching} parameterization against a HiFi4 reference
to verify that output tokens still match.

\paragraph{Fuse the MLP gate and up projections.} Merging the two
$3584 \!\to\! 5120$ projections into one $3584 \!\to\! 10240$ matmul
halves dispatch overhead and cuts one round-trip through activation
memory. We still need to check whether the stock config does this
already. If it does not, the fix is a small model-code edit followed by
another perf and accuracy run. This is the biggest remaining win after
TopK and HiFi2.

%% =========================================================================
\section{Roadblocks and Open Questions}
%% =========================================================================

\paragraph{No \texttt{tt-smi} on the box.} \texttt{tt-smi} is not
installed on the machine we have access to, so we read device health
off the runtime logs \texttt{tt-metal} prints on first program launch.
This works but is awkward. Installing \texttt{tt-smi} is on our list.

\paragraph{Compile-cache variance.} The two back-to-back runs of
architecturally identical models showed about a 44\% difference in
Matmul mean time per op. We are pretty sure this is compile-cache
state, since op counts and shapes are identical, but we want to confirm
it. For every optimization experiment going forward we will clear the
cache and report matched-cache deltas.

\paragraph{HiFi2 accuracy.} HiFi2 drops the lowest activation mantissa
bit. We have not yet confirmed that this leaves generation quality
intact on code prompts. We will use \texttt{ci-token-matching} and also
add a short code-completion sanity check.

\paragraph{Coder vs.\ Instruct.} Because Qwen2.5-Coder-7B is
architecturally identical to Qwen2.5-7B-Instruct, we cannot really
demonstrate ``Coder-ness'' at the kernel level. For the final report
we plan to include a small code-generation prompt set so that we can
show we are actually running the Coder weights, and not just the
generic 7B-Instruct model.

%% =========================================================================
\section{Next Steps}
%% =========================================================================
Between now and the final report, our plan is to land the TopK
multi-core fix and re-profile, then measure the HiFi2 MLP speedup and
check the token-match rate, then look at gate${+}$up fusion. Two more
things are on the list but have not been started yet: a single-device
(non-TP) run so we can separate out the CCL overhead from the rest of
the numbers, and a prefill-mode profile, since Blackhole's 512-token
prefill cutoff probably exposes a different set of hot paths than what
we see in decode.

%% =========================================================================
\balance
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
